---
title: "IS605Assignment5"
output: html_document
---
Problem Set 1
========================================

First, we will formulate functions that compute ATA and ATB

```{r}
ATA <- function(A){
  t(A) %*% A
}
ATB <- function(A,B){
  t(A) %*% B
}
A <- matrix(c(1,1,1,1,0,1,3,4),ncol=2)
B <- matrix(c(0,8,8,20),ncol=1)

ATA(A) # This shows a 2x2 matrix
ATB(A,B)
```

To solve for the least square, we will need to find ATB/ATA. To do this, we will need to find the inverse of ATA. Since ATA is a 2x2 matrix, I will need to use a helper function that will find the inverse of a 2x2 matrix. 

```{r}
inv2x2 <- function(x){
  inverse <- matrix(c(x[2,2],-1*x[2,1],-1*x[1,2],x[1,1]),ncol=2)/(x[1,1]*x[2,2]-x[1,2]*x[2,1]) 
  print(inverse) 
}
# We will build this helper function into our leastSquare function
leastSquare <- function(A,B){
  inv2x2(ATA(A)) %*% ATB(A,B) 
}
```
Our least square will be X = (1,4)
```{r}
leastSquare(A,B)
```
The squared error is equal to the square of B - A*(least square)
```{r}
(B - (A %*% matrix(c(1,4),ncol=1)))^2
```
If we exchange B for P, we will see that the error is zero:
```{r}
P <- matrix(c(1,5,13,17),ncol=1)
leastSquare(A,P)
```
Again, the least square will be X = (1,4), but the error will equal zero:
```{r}
(P - (A %*% matrix(c(1,4),ncol=1)))^2
```
We can easily demonstrate the error by subtracting the P vector from B
```{r}
B - P
```
This is the same as B minus A times the least square:
```{r}
B - (A %*% matrix(c(1,4),ncol=1))
```
Now we will show that the error is orthagonal to P and to each of the columnns in A. To do so, we will show 
the transpose of each of the column vectors in A multiplied by the error equals zero. Likewise for P.
```{r}
e <- B - P
t(P) %*% e
t(A[,1]) %*% e
t(A[,2]) %*% e
```
Problem Set 2
========================================
To avoid confusion with the variables in PS1, all of the variables for this next portion will be affixed with "2"
```{r}
data <- read.table("auto-mpg.data",header=F)
data <- as.matrix(data)
A2 <- matrix(data[,1:4],ncol=4)
B2 <- matrix(data[,5],ncol=1)
colnames(A2) <- c("displacement","horsepower","weight","acceleration")
colnames(B2) <- "mpg"
```
First, let's look at what dimensions ATA will take on, so we can best approach the calculation of its inverse:
```{r}
dim(ATA(A2))
```
Since it is a 4x4 matrix, we will use the myinverse function from Assignment 4 as a lemma:
```{r}
myinverse <- function(x){
  cofactorMatrix <- matrix(rep(0,length(x)),ncol=ncol(x))
  for (i in 1:ncol(x)) {
    for (j in 1:nrow(x)) {
      cofactorMatrix[i,j] <- det(x[-i,-j])*(-1)^(i+j) 
    }
  }
  print(t(cofactorMatrix)/det(x)) 
}
```
Now we will build a new least square function, using this inverse function:
```{r}
leastSquare2 <- function(A,B){
  myinverse(ATA(A)) %*% ATB(A,B) 
}

X2 <- leastSquare2(A2,B2)
```
The least square function should give us the variables that make up the variables that express the best fitting 
equation to calculate MPG in terms of the other 4 variables:
```{r}
X2
```
We can check these numbers with R's built in least square fit function:
```{r}
check <- lsfit(A2, B2, intercept=FALSE)
check$coefficients
```
The best approximation of mpg will be: 
 mpg = -0.030037938(displacement) + 0.157115685(horsepower) - 0.006217883(weight) + 1.997320955(acceleration)

The predicted error will be e = b - Ax
```{r}
e2 <- B2 - (A2 %*% X2)
head(e2)
```