---
title: "IS 605 Assignment 13"
author: "David Stern"
date: "May 2, 2015"
output: html_document
---

Part 1
------

For this exercise, we want to compute the derivative of $f(x) = x^{3} + 2x^{2}$ at any value of x. We will compute this approximation using limits. We know that for all values of $\textit{x}$ where the limit exists:

$$f'(x) = \lim_{h \to\ 0} \frac{f(x+h) - f(x)}{h}$$

We will build our function using this definition and a very small value of $\textit{h}$ to approximate the limit approaching zero. We will then round out the very small decimal at the end.

```{r}
derive <- function(x){
  result <- ((x+1e-5)^3 + 2*(x+1e-5)^2 - x^3 - 2*x^2)/1e-5
  result <- round(result,0)
  return(result)
}
```

We can easily compute the derivative of $f(x) = x^{3} + 2x^{2}$ as $f'(x) = 3x^{2} + 4x$. So we can test the derivative function simply by calling some small values:

$$f'(1)=7$$
$$f'(2)=20$$
$$f'(3)=39$$

```{r}
derive(1)
derive(2)
derive(3)
```

Now we will write a program that computes the area under the curve for $f(x) = 3x^{2} + 4x$ in the range $x = [1,3]. To approximate the area under the curve, we will sum the area of one million rectangles under the curve over the range.

First we create our curve:

```{r}
curve <- function(x){
 3*x^2 + 4*x
}
```

Now we will set $\Delta$ = 1e-6 and create a sequence of x-values corresponding to the left-border of each rectangle over the range x=[1,3].

```{r]}
deltaX <- 1e-6
AtoB <- seq(1,3,1e-6)
```

To compute the area, we sum the area of each rectangle:

```{r}
area <- 0
for (i in 1:length(AtoB)){
 area <- area + curve(AtoB[i])*deltaX
}
area
```

We can check our answer by calculating the definite integral:

$$\int_{1}^{3} 3x^{2} + 4x  dx = \left[x^{3} + 2x^{2}\right]_1^3 = (27 + 18) - (1 + 2) = 42$$

Part 2
------

Use integration by parts to solve for $\int sin(x)cos(x) dx$. For the following two problems, our formula will be :

$$\int fg'  dx = fg - \int f'g  dx$$

$\textbf{Problem 1}$: $\int sin(x)cos(x) dx$ For this problem, we will attempt to solve it with: $f(x) = g(x) = cos(x)$.

$$\int -sin(x)cos(x) dx = cos^{2}x - \int -sin(x)cos(x) dx$$
$$-\int sin(x)cos(x) dx = cos^{2}x + \int sin(x)cos(x) dx$$
$$-2\int sin(x)cos(x) dx = cos^{2}x$$
$$\int sin(x)cos(x) dx = -\frac{cos^{2}x}{2} + C$$

$\textbf{Problem 2}$: $\int x^{2}e^{x} dx$. For this problem, we will attempt to solve it with $f(x) = x^{2}$ and $g(x) = e^{x}$.

$$\int x^{2}e^{x} dx = x^{2}e^{x} - \int 2xe^{x} dx$$
$$\int x^{2}e^{x} dx = x^{2}e^{x} - 2\int xe^{x} dx$$
$$\textit{We will use integration by parts again to solve for} \int xe^{x}.$$
$$\int xe^{x} dx = xe^{x} - \int e^{x} dx$$
$$\int xe^{x} dx = xe^{x} - (e^{x} + C)$$
$$\textit{Now we will plug the formula for} \int xe^{x} \textit{into the equation}.$$
$$\int x^{2}e^{x} dx = x^{2}e^{x} - 2(xe^{x} - (e^{x} + C))$$
$$ = x^{2}e^{x} - 2xe^{x} + 2e^{x} + C$$
$$\textit{We will not double C alegbraically, as it is an arbitary constant.}$$

For the next problem, we will solve usingt the product rule of differentiation: $(fg)] = f'g + fg'$.

$\textbf{Problem 3}$:

$$\frac{d}{dx}(xcos(x)) = cos(x)\frac{d}{dx}(x) + x\frac{d}{dx}(cos(x))$$
$$ = cos(x) + x(-sin(x))$$
$$ = cos(x) - xsin(x)$$

$\textbf{Problem 4}$:

$$\frac{d}{dx}(e^{x^{4}}) = e^{x^{4}}\frac{d}{dx}(x^{4})$$
$$ = e^{x^{4}}4x^{3}$$

