---
title: "IS 607 Assignment 11"
author: "David Stern"
date: "April 20, 2015"
output: html_document
---

For this assignment, I will use the NYT Book Reviews API to retrieve information for books written by Christa Wolf.

```{r}
library(RCurl)
library(rjson)
```

When we call the JSON data, we will first transform it into a nested list. Data for each reviewed book has is stored in its own list, so we will run a double for loop to calleach piece of data from each listed book and store it in a more convenient format - a dataframe.

```{r}
callAPI <- getURL("http://api.nytimes.com/svc/books/v3/reviews.json?author=Christa%20Wolf&api-key=7d713de2214ccef5898be3e678f7ca2b:13:69774814")
jsonList <- fromJSON(callAPI) # Turn JSON data into a list
jsonList <- jsonList$results # Cut through the nested list and narrow down to the results list
christaWolf <- data.frame() #initialize dataframe
```

Run the loops:

```{r}
for (i in 1:length(jsonList)){
  for(j in 1:length(jsonList[[1]])){
    christaWolf[i,j] <- jsonList[[i]][[j]]
  }  
}
```

Now we will rename the columns of our dataframe:

```{r}
ourHeaders <- names(jsonList[[1]])
colnames(christaWolf) <- ourHeaders
christaWolf[,c(2,4,5,7)] # call part of data frame
```