---
title: "IS 661 Assignment 1"
author: "David Stern"
date: "February 24, 2016"
output: html_document
---

University of Minnesota's Introduction to Recommender Systems, Assignment 1.

Implementation in R:

```{r}
suppressWarnings(suppressMessages(require(operators)))
suppressWarnings(suppressMessages(require(plyr)))
movies <- read.csv("movies.csv",header=TRUE)
ratings <- read.csv("ratings.csv",header=TRUE)
```


Calculate, a movie's mean rating and return the top 10 movies by mean rating, with their mean, id number, and title.

```{r}
meanRatings <- aggregate(rating ~ movieId, ratings, mean)
joinedRatings <- join(meanRatings, movies, by = "movieId", type = "left")
reducedRatings <- joinedRatings[,3:1]
sortedRatings <- arrange(reducedRatings,-rating)
head(sortedRatings,10)
```

Calculate a movie's number of ratings and return the 10 movies with the most ratings, with their number of ratings, id number, and title.

```{r}
numRatingsTable <- table(ratings$movieId)
numRatingsDF <- as.data.frame(numRatingsTable)
colnames(numRatingsDF) <- c("movieId","numRatings")
numRatingsSorted <- arrange(numRatingsDF,-numRatings)
numRatingsSorted <- join(numRatingsSorted, movies, by="movieId", type="left")
numRatingsSorted <- numRatingsSorted[,-4] # shed genre column
head(numRatingsSorted,10)
```

Calculate a movie's damped mean, with a damping term of 5. Return the top 10 movies by damped mean rating, with their damped mean, id number, and title. 

$$\frac{\sum_{u}r_{ui}+k\mu}{n+k}$$
$$sum\:over\:users'\:ratings\:for\:an\:item$$
$$k\:= damping\:term$$
$$n\:= number of\:ratings$$
$$\mu\:=\:global\:mean$$

```{r}
globalMean <- mean(ratings$rating)
ratingSums <- aggregate(rating ~ movieId, ratings, sum)
ratingsByMovie <- join(ratingSums, numRatingsDF, by="movieId",type="left")
ratingsByMovie$dampedMean <- (ratingsByMovie$rating + 5*globalMean)/(ratingsByMovie$numRatings + 5)
ratingsByMovie <- arrange(ratingsByMovie,-dampedMean)
ratingsByMovie <- ratingsByMovie[,-2:-3] # drop columns
ratingsByMovie <- join(ratingsByMovie, movies, by="movieId", type="left") # add titles
ratingsByMovie <- ratingsByMovie[,-4] # drop genre column after join
head(ratingsByMovie,10)
```

Calculate the similarity of one movie to another based on how likely the user is to rate one given that they rated the other (ignoring the rating values), using the simple $\frac{x \wedge y}{x}$ method. Return the 10 movies most similar to movie 1389 (Jaws 3-D) using the simple metric, with their similarity scores, id number, and title.

```{r}
subRatings <- ratings[,1:2]
allUsers <- unique(subRatings$userId)
movie1389ratings <- subset(ratings,movieId==1389)
usersWhoRated1389 <- unique(movie1389ratings$userId)
subRatings$rated1389 <- 0
subRatings$didntRate1389 <- 0

for (i in 1:nrow(subRatings)){
  if (subRatings[i,1] %in% usersWhoRated1389) 
  {
    subRatings[i,3] <- 1
    subRatings[i,4] <- 0
  }
    else
  {
    subRatings[i,3] <- 0
    subRatings[i,4] <- 1
  }
}

head(subRatings)
```


```{r}
sumXY <- aggregate(cbind(rated1389,didntRate1389) ~ movieId, subRatings, sum)
sumXY$similarity1389 <- 0
sumXY$similarity1389 <- sumXY$rated1389/length(usersWhoRated1389)
sumXY <- arrange(sumXY,-rated1389)
sumXY <- join(sumXY, movies, by="movieId", type="left") 
sumXY <- sumXY[,c(4,5,1)] # drop unwanted columns
sumXY <- arrange(sumXY,-similarity1389)
head(sumXY,11)
```

Here we see that the most similar movie is Jaws 3D itself - a good indication that our formula worked.

Now we calculate the similarity of one movie to another, using the advanced $\frac{\frac{x \wedge y}{1+x}}{1+\frac{!x \wedge y}{1+!x}}$ method. We will return the 10 movies most similar to movie 1389 (Jaws 3-D) using the advanced metric, with their similarity scores, id number, and title.
 
First we need to recalculate sumXY with 1 in the denominator.

```{r}
sumXY <- aggregate(cbind(rated1389,didntRate1389) ~ movieId, subRatings, sum)
sumXY$similarity1389 <- 0
sumXY$dissimilarity1389 <- 0
sumXY$index <- 0
sumXY$similarity1389 <- sumXY$rated1389/(1+length(usersWhoRated1389))
sumXY$dissimilarity1389 <- (1+sumXY$didntRate1389)/(1+length(allUsers) - length(usersWhoRated1389))
sumXY$index <- sumXY$similarity1389/(1+sumXY$dissimilarity1389)

sumXY <- join(sumXY, movies, by="movieId", type="left") 
sumXY <- sumXY[,c(6,7,1)] # drop unwanted columns
sumXY <- arrange(sumXY,-index)
head(sumXY,11)
```