---
title: "IS 607 Project 2"
author: "David Stern"
output: html_document
---

In this exercise, we will do a analyze the data in four small datasets - each composed of 11 observations of two variables: x and y.

First, let's load the required packages:

```{r,echo=FALSE}
require(stringr)
require(plyr)
require(ggplot2)
require(grid)
require(gridExtra)
require(GGally)
require(scales)
require(reshape2)
```

And our data: 

```{r}
x1 <- c(10.0,8.0,13.0,9.0,11.0,14.0,6.0,4.0,12.0,7.0,5.0)
y1 <- c(8.04,6.95,7.58,8.81,8.33,9.96,7.24,4.26,10.84,4.82,5.68)
x2 <- c(10.0,8.0,13.0,9.0,11.0,14.0,6.0,4.0,12.0,7.0,5.0)
y2 <- c(9.14,8.14,8.74,8.77,9.26,8.10,6.13,3.10,9.13,7.26,4.74)
x3 <- c(10.0,8.0,13.0,9.0,11.0,14.0,6.0,4.0,12.0,7.0,5.0)
y3 <- c(7.46,6.77,12.74,7.11,7.81,8.84,6.08,5.39,8.15,6.42,5.73)
x4 <- c(8.0,8.0,8.0,8.0,8.0,8.0,8.0,19.0,8.0,8.0,8.0)
y4 <- c(6.58,5.76,7.71,8.84,8.47,7.04,5.25,12.50,5.56,7.91,6.89)
df1 <- data.frame(x1,y1) # Load each pair into a dataframe and then compile into a master dataframe
df2 <- data.frame(x2,y2)
df3 <- data.frame(x3,y3)
df4 <- data.frame(x4,y4)
master <- data.frame(df1,df2,df3,df4)
```

Let's get a brief overview of the stats. We can see that the mean and median are fairly stable among the pairs.

```{r}
#The following steps take the summary data and convert them into a "long" dataframe
sumStats <- as.data.frame(summary(master)) 
sumStats$Freq <- as.character(sumStats$Freq)
interlist <- str_split(sumStats$Freq, ":")
interdf <- ldply(interlist)
colnames(interdf) <- c("Stat", "Value")
sumStats <- cbind(sumStats, interdf)
sumStats <- sumStats[-c(1,3)]
colnames(sumStats)[1] <- "Variable"
sumStats
```

Now let's build some graphs to visualize the dataset.

```{r}
sumStats$Value <- as.numeric(sumStats$Value)
wideData <- reshape(sumStats, direction="wide", idvar="Variable", timevar="Stat")
wideData["Value.Range"] <- wideData$Value.Max - wideData$Value.Min # Let's add the Range and IQR of each set
wideData["Value.IQR"] <- wideData["Value.3rd Qu."] - wideData["Value.1st Qu."]
p1 <- ggplot(df1, aes(x=x1, y=y1)) + geom_line() # Now for visuals
p2 <- ggplot(df2, aes(x=x2, y=y2)) + geom_line()
p3 <- ggplot(df3, aes(x=x3, y=y3)) + geom_line()
p4 <- ggplot(df4, aes(x=x4, y=y4)) + geom_line()
grid.arrange(p1,p2,p3,p4,ncol=2,main="X,Y Correlations")
```

As we can see, the x and y have varying relationships in the datasets and the points in each dataframe seem
to be treated as ordered pairs. In the upper-left quadrant, df1 demonstrates a positive, but very volatile correlation. To its right, df2 looks to be parabolic - the x value is positively correlated with the y value until it reaches a peak. In the bottom-left quadrant, df3 demonstrates a steady positive correlation, with an outlier at (x3,y3) = (13,12.74). The following call will show the outlier in the raw data:

```{r}
df3[order(y3),]
```

Our bottom-right quadrant shows an odd distribution for df4. This set seems to have no correlation, as the x values remain the same for all values of y with the exception of one outlier. It's clearer if we look at the distribution with a scatterplot:

```{r}
ggplot(df4, aes(x=x4, y=y4)) + geom_point()
```

We can also look at a correlation of matrix of all 8 X and Y variables:

```{r}
cor(master[,1:8])
ggpairs(cor(master[,1:8]),axisLabels="internal")
```

The previous graph is very difficult to interpret. Instead, we will build a heatmap to visualize correlation.

```{r}
allCor <- cor(master[,1:8])
allCor2 <- melt(allCor,varnames=c("x","y"),value.name="Correlation")
allCor3 <- allCor2[order(allCor2$Correlation),]
ggplot(allCor3,aes(x=x, y=y)) +
  geom_tile(aes(fill=Correlation)) + 
  theme_minimal() +
  labs(x=NULL,y=NULL)
```

We can glean at least two useful pieces of information from this table:

1. The datasets df1,df2,df3 all have strong positive correlations with each other, but are all negatively correlated with df4.

2. The lightest tiles may indicate a near perfect correlation between the x variables in df1,df2, and df3. We will test the correlations with an equality.

```{r}
allCor["x1","x2"] == (allCor["x2","x3"] == allCor["x1","x3"]) # wrap the second pair to run multiple equalities
```