---
title: "IS 607 Assignment 9"
author: "David Stern"
date: "March 31, 2015"
output: html_document
---

In this exercise, we will attempt to pull identical data on three books out of three different file formats: JSON, HTML, and XML. We will create these file in R so we can see how each file is constructed. We will want to compile the data on each of the books into a dataframe.

```{r}
suppressWarnings(suppressMessages(library(rjson)))
suppressWarnings(suppressMessages(library(XML)))
```


This is the data we will want to encode:

| title                               | author                                | cuisine    | yearPub | ISBN       |
|-------------------------------------|---------------------------------------|------------|---------|------------|
| Plenty                              | Yolam Ottolenghi and Jonathan Lovekin | Vegetarian | 2011    | 1452101248 |
| Mastering the Art of French Cooking | Julia Child                           | French     | 1961    | 375413405  |
| Balaboosta                          | Einat Admony                          | Israeli    | 2013    | 1579655009 |

JSON
----

Here we will write our JSON data into a file in and see how R returns the data from the file.

```{r}
exportJSON <- "{\"title\":[\"Plenty\",\"Mastering the Art of French Cooking\",\"Balaboosta\"],\"author\":[\"Yolam Ottolenghi and Jonathan Lovekin\",\"Julia Child\",\"Einat Admony\"],\"cuisine\":[\"Vegetarian\",\"French\",\"Israeli\"],\"yearPub\":[2011,1961,2013],\"ISBN\":[1452101248,375413405,1579655009]}"
                
write(exportJSON, "cookBooks.json") # create JSON file
json_data <- fromJSON(file="cookBooks.json") # now we read directly from the file
json_data #view
```

Now we will compile this into a dataframe:

```{r}
jsonDF <- data.frame(json_data$title,json_data$author,json_data$cuisine,json_data$yearPub,json_data$ISBN)
for (i in 1:length(jsonDF)){colnames(jsonDF)[i] <- names(json_data)[i]}
```

HTML
----

First we will write our HTML data into a file and see how R returns the data:

```{r}
exportHTML <- "
<table>
  <tr>
  <th>title</th>
  <th>author</th>
  <th>cuisine</th>
  <th>yearPub</th>
  <th>ISBN</th>
  </tr>
  <tr>
  <td>Plenty</td>
  <td>Yolam Ottolenghi and Jonathan Lovekin</td>
  <td>Vegetarian</td>
  <td>2011</td>
  <td>1452101248</td>
  </tr>
  <tr>
  <td>Mastering the Art of French Cooking</td>
  <td>Julia Child</td>
  <td>French</td>
  <td>1961</td>
  <td>375413405</td>
  </tr>
  <tr>
  <td>Balaboost</td>
  <td>Einat Admony</td>
  <td>Israeli</td>
  <td>2013</td>
  <td>1579655009</td>
  </tr>
  </table>"
write(exportHTML, "cookBooks.html") # create HTML file
htmlDF <- readHTMLTable("cookBooks.html")
class(htmlDF) #returns as a list
htmlDF
htmlDF <- htmlDF$'NULL' # Now we have our data frame
```

XML
---

We will use the same tactic with XML as with the previous examples. Here, you will see the data is structured much differently. When we return the data, we will use the XML package to identify nodes and return specific data by its identifying tag. We will then manually construct the dataframe by author, title, cuisine type, publication year, and ISBN number.

```{r}
exportXML <- "
<myLibrary>
  <book>
      <title>Plenty</title>
      <author>Yolam Ottolenghi</author>
      <author>Jonathan Lovekin</author>
      <cuisine>Vegetarian</cuisine>
      <yearPub>2011</yearPub> 
      <ISBN>1452101248</ISBN>
  </book>
  <book>
      <title>Mastering the Art of French Cooking</title>
      <author>Julia Child</author>
      <cuisine>French</cuisine>
      <yearPub>1961</yearPub> 
      <ISBN>375413405</ISBN>
  </book>
  <book>
      <title>Balaboosta</title>
      <author>Einat Admony</author>
      <cuisine>Israeli</cuisine> 
      <yearPub>2013</yearPub>
      <ISBN>1579655009</ISBN>
  </book>
</myLibrary>"
  
write(exportXML, "cookBooks.xml") # create XML file

xmlData <- xmlTreeParse("cookBooks.xml",useInternal=T)
ourRoot <- xmlRoot(xmlData)
xmlName(ourRoot) # identify parent node
names(ourRoot) # find its child nodes - this shows there are three books in myLibrary
# like a list, calling top[1] will return the first book
ourRoot[1]
# we could identify the child nodes for each books by calling names(top[[1])
names(ourRoot[[1]])
# We can built this dataframe by pulling the values one at a time:
title <- xpathApply(xmlData, "//title", xmlValue)
title <- as.character(title)
author <- xpathApply(xmlData, "//author", xmlValue)
author <- as.character(author)
cuisine <- xpathApply(xmlData, "//cuisine", xmlValue)
cuisine <- as.character(cuisine)
yearPub <- xpathApply(xmlData, "//yearPub", xmlValue)
yearPub <- as.character(yearPub)
ISBN <- xpathApply(xmlData, "//ISBN", xmlValue)
ISBN <- as.character(ISBN)
```

Now we need to combine the two authors in the "author" vector for the book "Plenty". Since each row must represent one book, we need to make sure the two authors occupy one "cell".

```{r}
author[1] <- paste(author[1],"and",author[2])
author <- author[-2]
xmlDF <- data.frame(title,author,cuisine,yearPub,ISBN) # compile dataframe
```

Now we can see if the dataframes are identical:

```{r}
xmlDF
htmlDF
jsonDF
```



