---
title: "David Stern IS 606 Assignment 6"
output: html_document
---

The following exercise will create two R functions: 

allProbs - This function will take a file as input and return the probabilities of each word in the document.

detectTwo - This function will take a file and two words as input. It will return a matrix of the distinct probability of the occurence of each word as well as the adjacency of the two - the probability that the two words occur together, in either order.


```{r,tidy=TRUE}
allProbs <- function(x){
  text <- scan(x, character(0),quote=NULL)
  output <- tolower(gsub("[^[:alnum:][:space:]]", "", text))
  return(table(output)/length(output))
}
```
Now we can run the function by specifying the text file:

```{r}
P <- allProbs("/Users/davidstern/assign6.sample.txt")
head(P,100)
```

Let's make sure all the probabilities add up to 1:

```{r}
sum(allProbs("assign6.sample.txt"))
```

The following function will take a document and two words as input and return the probability of each of the words alone, as well as "adjacency" - the probability of them occuring together.

```{r,tidy=TRUE}
detectTwo <- function(doc,word1,word2){
  text <- scan(doc, character(0),quote=NULL) # Splits text into words
  output <- tolower(gsub("[^[:alnum:][:space:]]", "", text)) 
  #  ^ Removes punctuation and keeps alphanumeric characters and spaces
  totals <- table(output)/length(output) # stores probabilities of each word
  word1 <- as.character(word1) # converts word1/word2 inputs to character types
  word2 <- as.character(word2)
  count <- 0
  for(i in 1:length(text)){
    if (output[i] == word1 && (output[i-1] == word2 || output[i+1] == word2))
    { # This will count every instance where word2 precedes OR follows word1
      count <- count + 1
    }
  } # Now we will print our results in a tidy matrix
  results <- as.matrix(c(totals[word1],totals[word2],count/length(text)),byrow=TRUE)
  rownames(results) <- c(word1,word2,"adjacency")
  print(t(results))
}
```

Let's test the function with "the" and "for":

```{r}
detectTwo("/Users/davidstern/assign6.sample.txt","the","for")
```
